Emergent dual scaling of riverine biodiversity

### 2021-07-02 manuscript_pnas.Rmd

While the foundational concepts in ecology emphasize the importance of ecosystem size in biodiversity relationships [@hubbellUnifiedNeutralTheory2001; @triantisIslandSpeciesareaRelationship2012; @iwasakiEvaluatingRelationshipBasinscale2012; @saboRoleDischargeVariation2010], ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial [@turnerLandscapeEcologyTheory2015] and aquatic ecosystems [@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. In this context, this study substantially advanced our understanding of how biodiversity is organized in spatially complex ecosystems, serving as a conceptual pillar for future research. For example, the proposed framework may be extended to explore how these ecosystem properties regulate functional diversity [@petcheyFunctionalDiversityFD2002] and shifts in stochastic/deterministic community assembly [@chaseStochasticCommunityAssembly2010]. Recognizing the dual control of biodiversity broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our discovery should therefore bring a paradigm shift in macroecological science with far-reaching implications for biodiversity conservation.

>>>>

While the foundational concepts in ecology emphasize the importance of ecosystem size in biodiversity relationships [@hubbellUnifiedNeutralTheory2001; @triantisIslandSpeciesareaRelationship2012; @iwasakiEvaluatingRelationshipBasinscale2012; @saboRoleDischargeVariation2010], ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial [@turnerLandscapeEcologyTheory2015] and aquatic ecosystems [@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. In this context, the present study substantially advanced our understanding of how biodiversity is organized in spatially complex ecosystems, offering a conceptual pillar for future research. For example, the proposed framework may be extended to explore how these ecosystem properties regulate functional diversity [@petcheyFunctionalDiversityFD2002] and shifts in stochastic/deterministic community assembly [@chaseStochasticCommunityAssembly2010]. Recognizing the dual control of biodiversity broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our discovery should therefore bring a paradigm shift in macroecological science with far-reaching implications for biodiversity conservation.
